Group living is a fascinating game theoretic problem.
In many social contexts, the best overall outcome for all individuals occurs through universal or near-universal cooperation.
However, it is in the interest of each individual to defect.
Human social groups have developed sophisticated tools to protect the group from being undermined by defectors, key among these being discipline -- censuring selfish behavior and exiling irremediable defectors.
Wilson posits that a key horizontal effect of religion is to maintain discipline.
He points to the nuanced social codes prescribed by Calvanism, claiming that \blockcquote[p 250]{Wilson2007EvolutionLives}{religion can embody detailed instructions for\ldots retaliation leavened with just the right amount of generosity, contrition, and forgiveness conditioned upon a change of behavior}.
Calvanism originated in the environment of reformation-era Geneva, a large and sophisticated social group.
In contrast, the frontier is much less densely populated, playing host to posses of adventurers, family homesteads, and cow towns.
In fact, this population sparseness is the formal criteria that historian Frederick Turner relied on to define the frontier.
\blockcquote{Turner1894TheHistory}{The most significant thing about the American frontier is, that it lies at the hither edge of free land.
In the census reports it is treated as the margin of that settlement which has a density of two or more to the square mile}.
In the context of small or nonexistent social groups present in the frontier environment, simpler, more direct approaches to promoting group cohesion are practicable.
This fact is evident in McCarthy's \textit{Blood Meridian}.
In the midst of the wanderings of the kid's posse, two members defect from the group, abandoning their companions.
When the kid's posse encounters a camp set up by two defecting members, they wordlessly exact harsh retribution by destroying the defectors' gear.
\blockcquote[p 112]{McCarthy1992BloodWest}{He undid the girthstraps and loosed the other accouterments and piled them onto the flames, blankets, saddle, all, the greasy wool and leather sending up a foul gray smoke}.
No particular effort is made to justify or mask the directly retaliatory nature of this action; none is needed.
The frontier's harsh environ makes the cost to the group of individual defections -- a potentially violent or otherwise gruesome demise -- so stark that the vertical components of religion are not necessary to motivate or justify retaliatory disciplinary action.
