Although religion's horizontal effect of promoting group cohesion has been displaced, religion persists in \textit{Blood Meridian}.
So, what horizontal effects of religion remain relevant in McCarthy's frontier? Religion provides a framework for reconciling senseless hardships encountered by frontierspeople with an inherent human need for order and meaning.
A major vertical component of Christian religion is the belief that all events are part of an inscrutable overarching narrative and, further, and that individuals can attempt to transact through prayer for what they want but cannot have \cite[p 240]{Wilson2007EvolutionLives}.
The appeals to religion made by the dying man left behind at the old presido exemplify the horizontal functionality of frontier religion.
\blockcquote[p 119]{McCarthy1992BloodWest}{Within the doorless cuartel the man who'd been shot sang church hymns and cursed God alternately \ldots they could hear the hymns of their childhood \ldots}.
The dying man attempts to externalize his plight through imagined transactions with a higher power and by availing himself of long-gone vestiges of religious community via memories of the hymnal of organized religion, finding perhaps some cold comfort in these efforts.
Importantly, in this vignette religion does not activate scrupulous social behavior amongst the frontiersfolk. The kid's posse departs without raising \blockcquote[p 119]{McCarthy1992BloodWest}{a hand in farewell}.
Here, religion serves solely as an individual solace.

If the hypothesis of individual resilience as an element of horizontal religious functionality in the frontier were true, we would expect to observe magnification of aspects of the vertical dimension of religion that promote that horizontal functionality.
Indeed, in \textit{Blood Meridian} we observe a tailoring of the vertical dimension of frontier religion to accommodate these new horizontal ends in the theological ramblings of the judge.
\blockcquote[p 116]{McCarthy1992BloodWest}{Books lie, he said.
God dont lie.
No, said the judge.
He does not.
And these are his words.
He held up a chunk of rock.
He speaks in stones and trees, the bones of things}.
Here, the judge claims that no aspects of reality are arbitrary but are instead infused with a higher meaning we cannot hope to grasp but should respect.
The release from responsibility for distribution of fortune -- and therefore release from indignation at unfairness -- enabled by this doctrine is made explicit in dialogue Tobin delivers to the kid.
\blockcquote[p 123]{McCarthy1992BloodWest}{The gifts of the Almighty are weighed and parceled out in a ascale peculiar to himself.
It's no fair accountin and I dont doubt but what he'd be the first to admit it\ldots}.
Such a philosophy would indeed be useful to a forntiersperson fallen on hard times -- fortune is ultimately not the fault of themselves or fellow man but instead originates from a mysterious source impervious to human interference.
Thus, futile and exhausting emotional struggle against injustice is obviated.
As anticipated, the vertical component of frontier religion exhibits adaptation to promote horizontal functionality imperative to frontierspeople.
